---
title: "Analyzing Factors Associated with Area Burned by Wildfires in the United States"
author: "Info Innovators: Kevin Mao, Arnav Meduri, Ben Trokenheim, Ricardo Urena"
date: "April 28, 2025"
format:
  pdf:
    keep-tex: true
    number-sections: true
    fig-caption: true
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{table}{H}
        \usepackage{placeins}
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

## Introduction and Data

#### Background and Data Description

Wildfires are destructive natural disasters that occur regularly across the United States. According to the EPA, the U.S. has averaged approximately 70,000 wildfires per year since 1983[^1]. Although fire is a natural part of many ecosystems, wildfires have important economic and environmental consequences (e.g., property destruction, environmental degradation, and human health impacts). After recent wildfire incidents in California and western North Carolina, we became interested in better understanding the factors that contribute to the likelihood that a wildfire burns a greater-than-typical area and the factors that help explain variability in the continuous outcome of burned area. Consequently, identifying the factors associated with the extent of area burned can help inform decisions by wildfire management agencies. In light of this, we focused on two primary research questions: (a) What factors known before a wildfire has occurred are most strongly associated with the likelihood that a fire burns a greater-than-typical area? and (b) What overall factors (including those available after a wildfire) help explain variability in the continuous size of the burned area? To answer these research questions, we conducted exploratory data analysis and fitted both logistic and linear regression models to examine associations between wildfire characteristics and burned area.

[^1]: United States Environmental Protection Agency, "Climate Change Indicators: Wildfires," 2023, https://www.epa.gov/climate-indicators/climate-change-indicators-wildfires.

The dataset used in this analysis is an integrated dataset consisting of over 55,000 wildfire records from the United States between 1992 and 2015, compiled from the Fire Program Analysis system. In addition to wildfire-specific attributes recorded in this database, the dataset was supplemented with additional information from the Forest Service Research Data Archive, NOAA Integrated Surface Hourly Database, vegetation and land cover data from Meiyappan and Jain’s global land-use dataset, and geographic proximity data from SimpleMap’s World Cities Database. As part of our analysis, we used a subset of these variables, including fire size (measured in acres) as the response variable; cause of fire (categorized as missing/undefined, arson, debris burning, miscellaneous, campfire, fireworks, children, lightning, equipment use, smoking, railroad, structure, or powerline); temperature (°C), wind speed (meters per second), relative humidity (%), and precipitation (millimeters) recorded 30 days prior to the fire; containment time (measured as the number of days from discovery to containment); vegetation classification based on land cover (with categories Open Shrubland, Polar Desert/Rock/Ice, Secondary Tropical Evergreen Broadleaf Forest, Temperate Evergreen Needleleaf Forest, C3 Grassland/Steppe, Desert, and Water/Rivers); and remoteness (a unitless value between 0 and 1 representing the scaled distance from the nearest urban center).

#### Hypotheses

We hypothesize that (1) environmental conditions (e.g., temperature, humidity, wind, precipitation), geographic characteristics (e.g., remoteness, region), and vegetation type influence the likelihood that a wildfire burns a greater-than-typical area (with hotter, drier, windier conditions, greater remoteness, and more flammable vegetation expected to increase this likelihood), and (2) that conditions before and after discovery (including containment time and fire cause) explain variability in burned area, with human-related causes and adverse environmental and geographic factors associated with larger fires.

## Exploratory Data Analysis

### Data Cleaning

Before conducting our analysis, we applied many data cleaning steps to prepare our dataset for modeling and interpretation. One of the major decisions we made as part of our data cleaning process was to filter our response variable, acres burned, since the majority of observations in our dataset (over 13,000) recorded fires that burned one acre or less of land. Thus, we decided to focus only on the interquartile range (middle 50%) of wildfires by acres burned, since the goal of our analysis was to focus on wildfires that can reasonably be addressed during early containment efforts (rather than fires that had already expanded beyond an early intervention phase), and to restrict our analysis to a more practical range of fire sizes. In addition, we created a binary outcome based on fire size (i.e., grouping wildfires as either falling within the lower or upper 50% of fire sizes), which allowed us to focus on the likelihood of a wildfire exceeding the median size later on in our analysis. Our data cleaning process also involved transforming and consolidating variables to improve model interpretability and analysis. Since most precipitation variables were heavily left-skewed (i.e., most fires occurred without recent precipitation), we transformed all precipitation variables into binary indicators (0 = no precipitation, 1 = precipitation \> 0). Additionally, we grouped states into four broader regions (Northeast, Midwest, South, and West), vegetation types into environmental categories (e.g., Forest, Shrubland, Grassland), and fire causes into cause categories (e.g., Natural, Recreational, Infrastructure). We dropped all observations with missing values for variables of interest and converted categorical variables (e.g., vegetation type, fire cause, region) into factor variables to support modeling.

```{r}
#| label: load-packages-and-data
#| echo: false
#| warning: false

library(knitr)
library(lubridate)
library(stringr)
library(patchwork)
library(GGally)
library(tidyverse)
library(tidymodels)
library(rms)
library(tidyverse)
library(kableExtra)
library(Stat2Data)


wildfire <- read_csv("data/FW_Veg_Rem_Combined.csv")
```

```{r}
#| label: clean-data

wildfire_clean <- wildfire |>
  select(-(1:2)) |> 
  mutate(
    Prec_pre_30 = if_else(Prec_pre_30 == -1, NA_real_, if_else(Prec_pre_30 == 0, 0, 1)),
    Prec_pre_15 = if_else(Prec_pre_15 == -1, NA_real_, if_else(Prec_pre_15 == 0, 0, 1)),
    Prec_pre_7  = if_else(Prec_pre_7  == -1, NA_real_, if_else(Prec_pre_7  == 0, 0, 1)),
    Prec_cont   = if_else(Prec_cont   == -1, NA_real_, if_else(Prec_cont   == 0, 0, 1)),
    putout_time_num = parse_number(putout_time),
    Vegetation = factor(Vegetation),
    stat_cause_descr = factor(stat_cause_descr),
    state = factor(state),
    region = case_when(
      state %in% c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA") ~ "Northeast",
      state %in% c("IL", "IN", "MI", "OH", "WI", "IA", "KS", "MN", "MO", 
                   "NE", "ND", "SD") ~ "Midwest",
      state %in% c("DE", "FL", "GA", "MD", "NC", "SC", "VA", "DC", "WV", 
                   "AL", "KY", "MS", "TN", "AR", "LA", "OK", "TX") ~ "South",
      state %in% c("AZ", "CO", "ID", "MT", "NV", "NM", "UT", "WY", "AK", 
                   "CA", "HI", "OR", "WA") ~ "West"
    ) |> factor(),
    cause_category = case_when(
      stat_cause_descr == "Lightning" ~ "Natural",
      stat_cause_descr %in% c("Campfire", "Children", "Fireworks", "Smoking") ~ "Recreational",
      stat_cause_descr == "Debris Burning" ~ "Debris Burning",
      stat_cause_descr == "Arson" ~ "Arson",
      stat_cause_descr %in% c("Equipment Use", "Powerline", "Railroad", "Structure") ~ "Infrastructure",
      stat_cause_descr %in% c("Miscellaneous", "Missing/Undefined") ~ "Other/Unknown",
      TRUE ~ "Other"
    ) |> factor(),
    Vegetation_group = case_when(
      Vegetation %in% c(0, 9) ~ "Tundra/Barren",         
      Vegetation %in% c(4, 5, 6, 7) ~ "Grassland",        
      Vegetation %in% c(10, 11, 12) ~ "Shrubland",        
      Vegetation %in% c(13, 14, 15, 16) ~ "Forest",       
      Vegetation %in% c(1, 2, 3, 8) ~ "Wetland/Other",    
      TRUE ~ "Other"
    ) |> factor()
  ) |>
  filter(!is.na(region)) |> 
  drop_na(
    fire_mag, stat_cause_descr, Vegetation, remoteness, fire_size, putout_time,
    Prec_pre_15, Prec_pre_30, Prec_pre_7, Prec_cont,
    Hum_pre_15, Hum_pre_30, Hum_pre_7, Hum_cont,
    Temp_pre_15, Temp_pre_30, Temp_pre_7, Temp_cont,
    Wind_pre_15, Wind_pre_30, Wind_pre_7, Wind_cont
  ) |>
  filter(
    fire_size >= quantile(fire_size, 0.25, na.rm = TRUE),
    fire_size <= quantile(fire_size, 0.75, na.rm = TRUE)
  ) |>
  mutate(
  mid50_hi = factor(if_else(fire_size > median(fire_size, na.rm = TRUE), 1, 0), levels = c(0, 1))
)

write_csv(wildfire_clean, "data/wildfire_clean.csv")
```

#### Univariate EDA

As part of our EDA, we first examined the distribution of the response variable, fire size, to better understand its scale and variability. As mentioned previously, we focused on the middle 50% of wildfires by acres burned, which represent moderate-sized fires that are more likely to be responsive to early containment efforts. Based on the histogram of burned area (left panel), the distribution of moderate-sized wildfires appears right-skewed and unimodal. There is a clear peak at 3,978 observations corresponding to wildfires that burned less than 2 acres. Within this subset, the mean burned area is 11.84 acres, and the median is 5 acres. The middle 50% of burned areas falls between 2.5 acres (first quartile) and 14.7 acres (third quartile), with minimum and maximum values of 1.5 and 80 acres, respectively. The standard deviation is 15.62 acres, indicating substantial variability within this range. (Summary statistics for fire size are provided in the Appendix.) Many wildfires burned more than 30 acres, extending beyond the typical range and representing moderate outliers. For the binary fire size outcome (right panel), 54.2% of wildfires fall into the lower 50% (0) and 45.8% fall into the upper 50% (1), which reflects a roughly balanced distribution.

```{r}
#| label: fire-size-dist
#| echo: false
#| warning: false
#| fig-height: 3


p1 <- ggplot(wildfire_clean, aes(x = fire_size)) +
  geom_histogram(fill = "cadetblue4", color = "white", alpha = 0.8, binwidth = 2) +
  labs(
    title = "Histogram of Burned Area",
    x = "Burned Area (acres)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, hjust = 0),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9)
  )

p2 <- wildfire_clean |>
  ggplot(aes(x = mid50_hi, fill = as.factor(mid50_hi))) +
  geom_bar(color = "white", alpha = 0.8) +
  scale_fill_manual(
    values = c("0" = "cadetblue3", "1" = "cadetblue4"),
    labels = c("Lower 50%", "Upper 50%"),
    name = "Fire Size Group"
  ) +
  labs(
    title = "Distribution of Fire Size Group",
    x = "Fire Size Group",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, hjust = 0),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 9)
  )


(p1 | p2) + 
  plot_layout(widths = c(4, 1)) + 
  plot_annotation(
    title = "Distribution of Area Burned by Wildfires",
    subtitle = "Moderate-Sized Wildfires",
    theme = theme(
      plot.title = element_text(hjust = 0),
      plot.subtitle = element_text(hjust = 0)
    )
  )
```

Next, we visualized the distribution of the quantitative variables of interest in our dataset (temperature (°C), relative humidity (%), wind speed from 30 days prior to fire discovery (m/s), and remoteness). We first examined the distribution of temperature a month before discovery (top left panel), which appeared approximately normal and unimodal. (This is consistent with what we would expect, since temperature tends to change gradually across areas rather than clustering at specific values.) The median temperature for wildfire observations is 12.97 °C and the mean is 13.58 °C, with most observations between 6.53 °C and 21.36 °C. We then examined the distributions of wind speed thirty days before discovery and remoteness (top right and bottom left panels, respectively). Wind speed appears very slightly right-skewed and unimodal, while remoteness appears more strongly right-skewed and unimodal. This suggests that most fires occurred relatively close to populated areas (lower remoteness) and that most wind speeds were low to moderate. The median remoteness is 0.21 and the mean is 0.25, with most observations between 0.15 and 0.35. The median wind speed is 2.89 m/s and the mean is 2.87 m/s, with most observations between 2.06 m/s and 3.76 m/s. Lastly, we examined the distribution of relative humidity thirty days before discovery (bottom right panel), which appears left-skewed and unimodal. Lower humidity conditions were more common among wildfires in our data. The median relative humidity is 63.53% and the mean is 55.20%, with most observations between 48.80% and 69.97%. However, a large number of observations recorded a relative humidity of exactly 0%. Given that 0% relative humidity is unlikely under normal atmospheric conditions (since even very dry air typically contains some moisture), we decided to exclude these observations from our analysis.

```{r}
#| label: quantitative-univariate-eda
#| echo: false
#| warning: false

p1 <- ggplot(wildfire_clean, aes(x = Temp_pre_30)) +
  geom_histogram(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(
    title = "Temperature",
    subtitle = "30 days prior to fire discovery",
    x = "Temperature (°C)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
  plot.title = element_text(size = 10),         # make plot title smaller
  plot.subtitle = element_text(size = 8),       # make subtitle smaller
  axis.title.x = element_text(size = 8),         # x-axis label
  axis.title.y = element_text(size = 8),         # y-axis label
  axis.text = element_text(size = 7)             # axis tick numbers smaller
  )



p2 <- ggplot(wildfire_clean, aes(x = Hum_pre_30)) +
  geom_histogram(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(
    title = "Relative Humidity",
    subtitle = "30 days prior to fire discovery",
    x = "Relative Humidity (%)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
  plot.title = element_text(size = 10),         # make plot title smaller
  plot.subtitle = element_text(size = 8),       # make subtitle smaller
  axis.title.x = element_text(size = 8),         # x-axis label
  axis.title.y = element_text(size = 8),         # y-axis label
  axis.text = element_text(size = 7)             # axis tick numbers smaller
  )

p3 <- ggplot(wildfire_clean, aes(x = Wind_pre_30)) +
  geom_histogram(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(
    title = "Wind Speed",
    subtitle = "30 days prior to fire discovery",
    x = "Wind Speed (m/s)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
  plot.title = element_text(size = 10),         # make plot title smaller
  plot.subtitle = element_text(size = 8),       # make subtitle smaller
  axis.title.x = element_text(size = 8),         # x-axis label
  axis.title.y = element_text(size = 8),         # y-axis label
  axis.text = element_text(size = 7)             # axis tick numbers smaller
  )

p4 <- ggplot(wildfire_clean, aes(x = remoteness)) +
  geom_histogram(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(
    title = "Remoteness",
    subtitle = "Scaled distance from nearest urban area",
    x = "Remoteness (non-dimensional distance)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
  plot.title = element_text(size = 10),         # make plot title smaller
  plot.subtitle = element_text(size = 8),       # make subtitle smaller
  axis.title.x = element_text(size = 8),         # x-axis label
  axis.title.y = element_text(size = 8),         # y-axis label
  axis.text = element_text(size = 7)             # axis tick numbers smaller
  )

(p1 | p3) / (p4 | p2) + 
  plot_annotation(
    title = "Distribution of Quantitative Environmental Variables",
    theme = theme(
      plot.title = element_text(size = 11)
    )
  )
```

In addition to examining the distributions of quantitative predictors, we also visualized the distribution of categorical variables of interest in our dataset (i.e., wildfire cause category, vegetation group, region, and precipitation 30 days prior to wildfire discovery), which can be found in the Appendix.

\FloatBarrier

#### Bivariate EDA

```{r}
#| label: fire-size-bivariate
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| out.width: "90%"

sp1 <- ggplot(wildfire_clean, aes(x = Temp_pre_30, y = fire_size)) +
  geom_point(alpha = 0.3, color = "cadetblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "gray40") +
  labs(
    title = "Temperature vs. Fire Size",
    subtitle = "30 days prior to fire discovery",
    x = "Temperature (°C)",
    y = "Burned Area (acres)"
  ) +
  theme_minimal()

sp2 <- ggplot(wildfire_clean, aes(x = Hum_pre_30, y = fire_size)) +
  geom_point(alpha = 0.2, color = "cadetblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "gray40") +
  labs(
    title = "Relative Humidity vs. Fire Size",
    subtitle = "30 days prior to fire discovery",
    x = "Relative Humidity (%)",
    y = "Burned Area (acres)"
  ) +
  theme_minimal()

sp3 <- ggplot(wildfire_clean, aes(x = Wind_pre_30, y = fire_size)) +
  geom_point(alpha = 0.2, color = "cadetblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "gray40") +
  labs(
    title = "Wind Speed vs. Fire Size",
    subtitle = "30 days prior to fire discovery",
    x = "Wind Speed (m/s)",
    y = "Burned Area (acres)"
  ) +
  theme_minimal()

sp4 <- ggplot(wildfire_clean, aes(x = remoteness, y = fire_size)) +
  geom_point(alpha = 0.2, color = "cadetblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "gray40") +
  labs(
    title = "Remoteness vs. Fire Size",
    subtitle = "Scaled distance from nearest urban area",
    x = "Remoteness (non-dimensional distance)",
    y = "Burned Area (acres)"
  ) +
  theme_minimal()

(sp1 | sp2) /
(sp3 | sp4) +
  plot_annotation(
    title = "Relationship Between Environmental Factors and Burned Area",
    subtitle = "Quantitative Variables", 
    theme = theme(plot.title = element_text(hjust = 0))
  )

```

As the next step in our analysis, we examined the relationship between each of our quantitative predictors and fire size to better understand their associations. In the scatterplot of burned area versus temperature (top left panel), the relationship appears very weak and slightly positive (i.e., larger fires may be slightly more likely at higher temperatures), but nonlinear, since most fires were small across the temperature range, and the largest fires (over 60 acres) tended to occur between 5 °C and 20 °C. The scatterplot of relative humidity versus burned area (top right panel) also shows a very weak, highly nonlinear, and slightly negative relationship, with fires of all sizes occurring across the full humidity range and the largest fires more common between 25% and 75% humidity. For wind speed (bottom left panel), the relationship appears very weak, slightly positive, and nonlinear; most fires occurred at lower wind speeds, and most larger fires (over 40 acres) were concentrated below 5 m/s, suggesting that higher wind speeds were not consistently associated with larger burned areas. In the scatterplot of remoteness versus burned area (bottom right panel), we observe a weak, slightly positive, and nonlinear relationship, with the largest fires occurring between remoteness values of 0.25 and 0.50 and relatively few large fires at very high remoteness values above 0.75. Overall, the wide vertical spread of data points around the lines of best fit for all predictors suggests that these quantitative predictors have limited predictive power for burned area when considered individually, which is expected given that wildfire size is influenced by many factors beyond the variables available in our dataset. (Summary statistics and containment time EDA are provided in the Appendix.)

```{r}
#| label: categorical-bivariate-dist
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

wildfire_clean <- wildfire_clean |>
  mutate(
    Precipitation = factor(
      if_else(Prec_pre_30 == 1, "Precipitation", "No Precipitation"),
      levels = c("No Precipitation", "Precipitation")
    )
  )

wildfire_clean_long <- wildfire_clean |>
  pivot_longer(
    cols = c(Vegetation_group, cause_category, region, Precipitation),  
    names_to = "Group_Type",
    values_to = "Group"
  ) |>
  mutate(
    Group_Type = dplyr::case_match(
      Group_Type,
      "Vegetation_group" ~ "Vegetation Type",
      "cause_category"   ~ "Cause Category",
      "region"           ~ "Region",
      "Precipitation"    ~ "Precipitation",
      .default = Group_Type
    )
  )

ggplot(data = wildfire_clean_long, 
       aes(x = Group, y = fire_size, color = Group)) +
  geom_boxplot(fill = NA, outlier.alpha = 0.6) +
  coord_flip() +
  facet_wrap(~ Group_Type, scales = "free_y") +
  scale_color_manual(
    values = colorRampPalette(c("#8FBC8F", "cadetblue4"))(
      length(unique(wildfire_clean_long$Group))
    )
  ) +
  labs(
    title = "Relationship Between Environmental Factors and Burned Area",
    subtitle = "Categorical Variables",
    x = NULL,
    y = "Burned Area (acres)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0, size = 12),
    plot.subtitle = element_text(hjust = 0, size = 10),
    legend.position = "none"
  )
```

As part of our bivariate EDA, we also analyzed the relationship between each of our categorical variables of interest and burned area to better understand differences in fire size across groups. We can see from the visualization (top left panel) that the distribution of acres burned is relatively similar for all wildfire cause categories, with medians between 4.00 and 6.00 acres, means between 9.02 and 13.74 acres, and outliers beyond the 30-acre mark across all cause types. This indicates that cause category on its own does not explain much variability in fire size. We observed a similar pattern between observations corresponding to no precipitation 30 days before discovery and observations corresponding to precipitation (top right panel), with both groups showing medians of 5.00 acres and means of 12.32 acres (no precipitation) and 11.38 acres (precipitation), although there was slightly more spread in the middle 50% of fires with no precipitation. We observed more apparent differences in the distribution of acres burned across different regions and vegetation groups (bottom left and bottom right panels, respectively). In particular, the Northeast had a lower median (3.00 acres) and mean (6.04 acres) compared to other regions, with much less variability in the middle 50% of fires and most outliers limited to about the 40-acre mark. In contrast, distributions for the West, South, and Midwest were more similar, with medians between 5.00 and 6.00 acres, means between 12.00 and 13.00 acres, and many outliers beyond the 30-acre mark. For vegetation type, distributions were fairly similar across categories, although grasslands had a slightly lower median (4.00 acres) and mean (8.43 acres) compared to other vegetation groups, with somewhat fewer extreme outliers. Overall, region and vegetation group seem to explain more variability in fire size compared to cause category or precipitation, and could be useful predictors in modeling fire size.

#### Potential Interaction Effects

Since many of our variables of interest relate to environmental conditions and geographic location, there are potential interaction effects among our predictors that could influence acres burned. As part of our analysis, we were interested in a potential interaction between vegetation type and precipitation, since vegetation and fuel moisture could influence burned area together. Through our EDA, we observed that acres burned did not seem to vary much across different precipitation groups within each vegetation type. (A more detailed analysis of this pattern is provided in the Appendix.)

## Methodology

The main goal of our analysis was to examine how various wildfire-related factors are related to burned area, and how these factors can be used to predict fire size. Since these represent two distinct research questions, we chose to fit two separate models: a linear regression model using all variables of interest (including those only available after a fire is contained) to explain variability in burned area (i.e., an explanatory model), and a logistic regression model using only variables that are available at or before fire discovery to guide response efforts to predict whether a wildfire burns a greater-than-typical area (i.e., a predictive model). (In this setup, the model estimates the likelihood that a wildfire falls into the upper half of moderate-sized fires based on available environmental factors.)

#### Multiple Linear Regression

To guide predictor selection for the explanatory model, we used a forward selection approach based on adjusted $R^2$, ultimately including all predictors of interest (each contributed to an increase in adjusted $R^2$). Our initial multiple linear regression predicted burned area based on region, cause category, remoteness, vegetation group, temperature, precipitation, wind speed, relative humidity (15 days before discovery), containment time, and an interaction between vegetation group and precipitation. We mean-centered the quantitative predictors to improve coefficient interpretability. After fitting this model, we conducted diagnostics to assess linearity, normality, constant variance, and independence. Normality, linearity, and independence appeared reasonably satisfied, but constant variance was violated. To address constant variance, we applied a variance-stabilizing (logarithmic) transformation to the fire size variable. We also found concerning multicollinearity between levels corresponding to the Southern and Western regions, which we addressed by combining these levels into a single group. A full assessment of model fit, including additional diagnostic checks, is provided in the Appendix.

#### Logistic Regression

To guide variable selection for the predictive model, we fit a full model including all variables of interest. This approach was appropriate because our dataset contained a limited number of predictors corresponding to information available prior to fire discovery, making it practical to include all relevant variables rather than selecting a smaller subset. Specifically, our initial logistic regression model predicted the likelihood that a wildfire would burn a greater-than-typical area based on temperature, humidity, wind speed, precipitation, vegetation type, the interaction between vegetation type and precipitation, remoteness, and region. After fitting the model, we determined that the key assumptions for logistic regression (linearity, randomness, and independence) were satisfied. A detailed assessment of model diagnostics (i.e., assessment of logistic regression conditions) can be found in the Appendix.

## Results

#### Multiple Linear Regression

```{r}
#| results: asis
exp_fire_fit <- lm(fire_size ~ region + cause_category + remoteness + Vegetation_group + 
                   Wind_pre_15 + Prec_pre_15 + Temp_pre_15 + Hum_pre_15 , data = wildfire_clean)

wildfire_clean <- wildfire_clean |>
  mutate(
    Prec_pre_15_cent = scale(Prec_pre_15, center = TRUE, scale = FALSE),
    Wind_pre_15_cent = scale(Wind_pre_15, center = TRUE, scale = FALSE),
    Temp_pre_15_cent = scale(Temp_pre_15, center = TRUE, scale = FALSE),
    Hum_pre_15_cent = scale(Hum_pre_15, center = TRUE, scale = FALSE)
  )

wildfire_clean$region <- recode(wildfire_clean$region, 
                                "South" = "South_West", 
                                "West" = "South_West")

fire_linear_fit <- lm(log(fire_size) ~ region + cause_category + remoteness + 
                               Vegetation_group * Prec_pre_15_cent + Wind_pre_15_cent + Temp_pre_15_cent +
                                 Hum_pre_15_cent + putout_time_num,
                               data = wildfire_clean)

tidy(fire_linear_fit, conf.int = TRUE) |>
 kable(digits = 3, format = "latex", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, font_size = 7, latex_options = c("hold_position"))
```

Our final multiple linear regression model predicting log-transformed fire size achieved an $R^{2}$ of 0.052 and an adjusted $R^2$ of 0.050 (model results in Appendix). This indicates that the model explains approximately 5.0% of the variability in log-transformed burned area among moderate-sized wildfires in the dataset. While low, this amount of explained variance is expected given the variability in wildfire behavior and the limited set of predictors available. We examined several predictors in our model and found that region, cause category, remoteness, vegetation group, wind speed 15 days prior to discovery, temperature 15 days prior to discovery, putout time, and the interaction between shrubland vegetation and precipitation were significant predictors of log-transformed fire size (p \< 0.05). In contrast, relative humidity 15 days prior (p \> 0.05), precipitation 15 days prior (p \> 0.05), and the interaction terms between grassland or tundra/barren vegetation and precipitation (p \> 0.05) were not significant; this result is consistent with our EDA findings (i.e., differences across region and vegetation type and the effect of wind speed were more apparent, while precipitation and humidity did not show strong relationships with fire size). The intercept of the model is estimated at 2.000, indicating that the expected fire size is approximately $e^{2.000} \approx 7.39$ acres when all predictors are at their reference or mean-centered values (i.e., for a wildfire in the Midwest region, caused by arson, in a forested area, under average environmental conditions — mean temperature = 13.8°C, mean relative humidity = 52.0%, mean wind speed = 2.77 m/s, mean remoteness = 0.252 — and with putout time equal to 0). Fires caused by recreational activities are expected to burn about $e^{-0.370} \approx 0.69$ times the acres compared to fires caused by arson, holding other factors constant. Fires caused by debris burning are expected to burn approximately $e^{-0.255} \approx 0.78$ times the acres compared to arson-related fires. Additionally, fires occurring in the Northeast are expected to burn about $e^{-0.942} \approx 0.39$ times the acres relative to fires in the Midwest. Other predictors, including vegetation group, environmental conditions (such as wind speed and temperature), remoteness, containment time, and vegetation–precipitation interactions, also showed statistically significant associations with fire size, although their effects were generally smaller in magnitude.

#### Logistic Regression

```{r}
#| label: logistic-model
#| results: asis

fire_log_fit <- glm(mid50_hi ~ region + remoteness + Vegetation_group * Prec_pre_15 + 
                   Wind_pre_15 + Temp_pre_15 + Hum_pre_15, data = wildfire_clean,
                   family = "binomial")

tidy(fire_log_fit, conf.int = TRUE) |>
  kable(digits = 3, format = "latex", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, font_size = 7, latex_options = c("hold_position"))
```

Our final logistic regression model predicting whether a wildfire burns a greater-than-typical area achieved an AUC of 0.569, indicating a relatively limited ability to distinguish between smaller and larger moderate-sized fires based on the predictors included (AUC value and corresponding ROC curve are provided in the Appendix). After fitting the full model with an interaction between vegetation group and precipitation, we conducted a drop-in-deviance test to assess whether the interaction terms were jointly useful. Since the p-value for the test was 0.013, we retained the interaction between vegetation group and precipitation (full test results are provided in the Appendix); while the individual interaction terms were not significant at the 0.05 level, the combined contribution significantly improved model fit. We examined several predictors in our model and found that region, remoteness, vegetation group (grassland and shrubland), wind speed 15 days prior, and temperature 15 days prior were significant predictors of fire size classification (p \< 0.05). In contrast, relative humidity 15 days prior (p \> 0.05), precipitation 15 days prior (p \> 0.05), tundra/barren vegetation group (p \> 0.05), and the interaction terms between vegetation group and precipitation (p \> 0.05) were not significant. The intercept of the model is estimated at -0.220, which indicates that the expected odds of a wildfire falling into the larger half of moderate-sized fire sizes is approximately $e^{-0.220} \approx 0.80$ when all predictors are at their reference or mean-centered values (i.e., for a wildfire in the Midwest region, in a forested area, with no precipitation 15 days prior, and under average environmental conditions — mean temperature = 13.8°C, mean relative humidity = 52.0%, mean wind speed = 2.77 m/s, mean remoteness = 0.252). Fires in the Northeast are expected to have approximately $e^{-1.830} \approx 0.16$ times the odds of falling into the upper 50% compared to fires in the Midwest, holding all else constant. Fires occurring in shrublands are associated with approximately $e^{0.209} \approx 1.23$ times the odds compared to fires occurring in forests, controlling for other variables. Other predictors such as remoteness, wind speed, and temperature also had statistically significant effects on fire size classification, although their individual impacts were smaller in magnitude.

## Discussion and Conclusion

Our analysis used both multiple linear regression and logistic regression to understand factors associated with wildfire size in the United States. The multiple linear regression model, which predicted log-transformed fire size, found that region, cause category, remoteness, vegetation group, wind speed 15 days prior to discovery, temperature 15 days prior to discovery, putout time, and the interaction between shrubland vegetation and precipitation were significant predictors (p \< 0.05), though relative humidity, precipitation, and interactions involving grassland and tundra/barren vegetation were not. The model explained approximately 5.0% of the variability in log-transformed fire size ($R^2 = 0.052$, adjusted $R^2 = 0.050$), which was expected given the variability in wildfire behavior and the predictors available. The logistic regression model, which predicted the likelihood that a wildfire burned a greater-than-typical area, used similar predictors except cause category, and achieved an AUC of 0.569, indicating limited classification ability. In both models, fires in shrubland and grassland were associated with larger burned areas, while fires in the Northeast tended to be smaller. Greater wind speeds slightly increased fire size, and greater remoteness was associated with smaller fires.

Our analysis had some limitations. Although the dataset included a large number of fires, it lacked information on factors that could explain additional variability in fire size (e.g., fuel characteristics, suppression resources, and local management). Because the data spanned more than two decades, unmodeled changes in climate, land use, and fire management may have introduced variability that the models did not capture. Although most model assumptions were reasonably satisfied after transformations, constant variance issues remained and could affect the reliability of coefficient estimates. Future work could focus on improving data coverage by collecting more detailed environmental and situational information for each fire. Although collecting data in remote regions can be challenging, more complete datasets could improve understanding of fire behavior and support resource allocation for fire management.

\newpage

## Appendix

#### Exploratory Data Analysis - Univariate EDA for Response Variable

A table of summary statistics for our response variable, fire size, and the binary fire size outcome (mid-50% classification) is included below.

```{r}
#| label: fire-size-summary-stats
#| echo: false
#| warning: false

table1 <- wildfire_clean |>
  summarise(
    Count = n(),
    Min = min(fire_size, na.rm = TRUE),
    Q1 = quantile(fire_size, 0.25, na.rm = TRUE),
    Median = median(fire_size, na.rm = TRUE),
    Mean = mean(fire_size, na.rm = TRUE),
    Q3 = quantile(fire_size, 0.75, na.rm = TRUE),
    Max = max(fire_size, na.rm = TRUE),
    SD = sd(fire_size, na.rm = TRUE)
  )

table2 <- wildfire_clean |>
  group_by(mid50_hi) |>
  summarise(
    Count = n(),
    Proportion = n() / nrow(wildfire_clean)
  ) |>
  mutate(Proportion = scales::percent(Proportion)) |>
  rename(
    "Fire Size Group" = mid50_hi,
    "Count" = Count,
    "Proportion" = Proportion
  )

kable(table1, digits = 2, format = "latex", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, font_size = 7)

kable(table2, digits = 2, format = "latex", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, font_size = 7)
```

#### Exploratory Data Analysis - Univariate EDA for Containment Time

We used putout time (containment time) as a predictor in our multiple linear regression model. As we can see from the plot below, the distribution of containment time is highly right-skewed, with most fires contained on the same day they were discovered (median = 0 days, mean = 0.98 days) and a small number of fires with much longer containment times (maximum = 261 days).

```{r}
#| fig-width: 8

ggplot(wildfire_clean, aes(x = as.factor(putout_time_num))) +
  geom_bar(fill = "cadetblue4", color = "white", alpha = 0.8, width = 0.8) +  
  labs(
    title = "Distribution of Containment Time",
    subtitle = "Containment time (in days)",
    x = "Containment Time (days)",
    y = "Count"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 90, check.overlap = TRUE)) +  # rotate AND avoid label collisions
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 7)
  )
```

#### Exploratory Data Analysis - Univariate EDA for Categorical Predictors

Looking more closely at vegetation group, we can see that most fires occurred in forested areas (5181 observations or 48.6%), followed by shrubland (2711 observations or 25.4%) and tundra or barren areas (2590 observations or 24.3%), with relatively few fires in grassland environments (189 observations or 1.77%). Examining wildfire cause categories, we found that a large portion of fires were attributed to human-related causes rather than natural causes. Debris burning accounted for the highest proportion (2821 observations or 26.4%), followed by arson (2399 observations or 22.5%) and other or unknown causes (2228 observations or 20.9%). Natural causes made up a smaller proportion (1535 observations or 14.4%), followed by recreational causes (853 observations or 8.0%) and infrastructure-related causes (835 observations or 7.82%). Finally, we examined the regional distribution of fires. Most fires in our dataset occurred in the South (6090 observations or 57.1%), followed by the West (2801 observations or 26.2%), Midwest (1228 observations or 11.5%), and Northeast (550 observations or 5.15%). Only two fires (2 observations or less than 0.01%) were categorized as occurring in the "Other" region classification.

```{r}
#| label: categorical-dist
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

base_theme <- theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 <- ggplot(wildfire_clean, aes(x = cause_category)) +
  geom_bar(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(title = "Wildfire Cause Category", x = NULL, y = "Count") +
  base_theme

p2 <- ggplot(wildfire_clean, aes(x = Vegetation_group)) +
  geom_bar(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(title = "Vegetation Group", x = NULL, y = "Count") +
  base_theme

p3 <- ggplot(wildfire_clean, aes(x = region)) +
  geom_bar(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(title = "Region", x = NULL, y = "Count") +
  base_theme

p4 <- ggplot(wildfire_clean, aes(x = Precipitation)) +
  geom_bar(fill = "cadetblue4", color = "white", alpha = 0.8) +
  labs(title = "Precipitation", x = NULL, y = "Count") +
  base_theme

# Combine with normal patchwork
combined_plot <- (p1 | p2) /
                 (p3 | p4) +
  plot_annotation(
    title = "Distribution of Categorical Environmental Variables"
  )

combined_plot

categorical_vars <- c("cause_category", "Vegetation_group", "region", "Precipitation")
```

#### Exploratory Data Analysis - Univariate EDA for Quantitative Predictors

A table of summary statistics for the quantitative predictors considered during modeling (temperature, relative humidity, wind speed, and remoteness) is included below.

```{r}
#| label: quantitative-summary-stats
#| echo: false
#| warning: false

wildfire_clean |>
  summarise(
    Temperature_Min = min(Temp_pre_30, na.rm = TRUE),
    Temperature_Q1 = quantile(Temp_pre_30, 0.25, na.rm = TRUE),
    Temperature_Median = median(Temp_pre_30, na.rm = TRUE),
    Temperature_Mean = mean(Temp_pre_30, na.rm = TRUE),
    Temperature_Q3 = quantile(Temp_pre_30, 0.75, na.rm = TRUE),
    Temperature_Max = max(Temp_pre_30, na.rm = TRUE),
    Temperature_SD = sd(Temp_pre_30, na.rm = TRUE),

    Humidity_Min = min(Hum_pre_30, na.rm = TRUE),
    Humidity_Q1 = quantile(Hum_pre_30, 0.25, na.rm = TRUE),
    Humidity_Median = median(Hum_pre_30, na.rm = TRUE),
    Humidity_Mean = mean(Hum_pre_30, na.rm = TRUE),
    Humidity_Q3 = quantile(Hum_pre_30, 0.75, na.rm = TRUE),
    Humidity_Max = max(Hum_pre_30, na.rm = TRUE),
    Humidity_SD = sd(Hum_pre_30, na.rm = TRUE),

    Wind_Min = min(Wind_pre_30, na.rm = TRUE),
    Wind_Q1 = quantile(Wind_pre_30, 0.25, na.rm = TRUE),
    Wind_Median = median(Wind_pre_30, na.rm = TRUE),
    Wind_Mean = mean(Wind_pre_30, na.rm = TRUE),
    Wind_Q3 = quantile(Wind_pre_30, 0.75, na.rm = TRUE),
    Wind_Max = max(Wind_pre_30, na.rm = TRUE),
    Wind_SD = sd(Wind_pre_30, na.rm = TRUE),

    Remoteness_Min = min(remoteness, na.rm = TRUE),
    Remoteness_Q1 = quantile(remoteness, 0.25, na.rm = TRUE),
    Remoteness_Median = median(remoteness, na.rm = TRUE),
    Remoteness_Mean = mean(remoteness, na.rm = TRUE),
    Remoteness_Q3 = quantile(remoteness, 0.75, na.rm = TRUE),
    Remoteness_Max = max(remoteness, na.rm = TRUE),
    Remoteness_SD = sd(remoteness, na.rm = TRUE)
  ) |>
  pivot_longer(everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_") |>
  pivot_wider(names_from = Statistic, values_from = value) |>
  kable(digits = 2) |>
  kable_styling(full_width = FALSE, font_size = 7, latex_options = c("hold_position"))
```

#### Exploratory Data Analysis - Potential Interaction Effects

In our analysis, we were interested in a potential interaction between vegetation type and precipitation. Based on the visualization below, the distributions of burned area appearrelatively similar between precipitation and no precipitation conditions for each vegetationgroup. In forest, the median burned area is 5.0 acres for both no precipitation (mean of 11.48acres) and precipitation (mean of 11.14 acres), while in grassland, the median is 3.5 acresfor no precipitation (mean 12.26 acres) and 3.0 acres for precipitation (mean of 11.54 acres).Additionally, in shrubland, the median is 6.0 acres for no precipitation (mean of 13.52 acres)and 5.0 acres for precipitation (mean of 11.31 acres), while in tundra/barren, the median is 5.0 acres for both no precipitation (mean of 12.81 acres) and precipitation (mean of 11.86 acres).Even though there does not appear to be a very strong effect, this interaction would be worthexploring more formally in the modeling process to determine whether it significantly improvesmodel fit.

```{r}
#| label: veg-prec-int
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

ggplot(wildfire_clean, aes(x = Vegetation_group, y = fire_size, color = Precipitation)) +
  geom_boxplot(fill = NA, outlier.alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Relationship Between Vegetation Type and Precipitation",
    subtitle = "Grouped by Burned Area",
    x = "Vegetation Group",
    y = "Burned Area (acres)",
    color = "Precipitation"
  ) +
  theme(
    plot.title = element_text(size = 11, hjust = 0),
    plot.subtitle = element_text(size = 9, hjust = 0),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9)
  ) +
  scale_color_manual(values = c("No Precipitation" = "cadetblue4", "Precipitation" = "#8FBC8F")) +
  coord_flip()
```

#### Methods - Explanatory Model

After fitting our initial linear regression model, we conducted a full set of diagnostics to assess linearity, normality, constant variance, and independence. A scatterplot of residuals vs. fitted values for our explanatory model is shown below. Given that the initial model we fit had a clear violation of constant variance (an increase in the variability of residuals as fitted values increased), we applied a variance-stabilizing (i.e., logarithmic) transformation to the response variable. We observed an improvement in the spread of residuals (more uniform spread across the range of fitted values) as well as a slight improvement in model fit (an increase in adjusted $R^2$) after applying this transformation. Because there was also a clear pattern in the residuals (i.e., they were not randomly scattered around $residuals = 0$), we examined scatterplots of fitted values versus each of the predictors and observed no clear nonlinear trends. As a result, we did not transform our predictors. We also did not observe any influential points ($D_i > 0.5$) in our data, so no observations were removed. Lastly, we observed two distinct groupings in the scatterplot of residuals vs. fitted values (one group with fitted values less than 1.5 and one group with fitted values generally above 1.5). We found that a large portion of this separation could be explained by region, so we fit separate linear regression models for different regional groups. However, this approach did not lead to a meaningful increase in explanatory power, so we retained a single model for all observations. From our work, we deem that constant variance and linearity are sufficiently satisfied.

```{r}
#| label: residuals-fitted-vals-plot
#| fig-width: 8
#| figh-height: 5

fire_linear_fit_aug <- augment(fire_linear_fit)

ggplot(fire_linear_fit_aug, aes(x = .fitted, y = .std.resid, color = .cooksd)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Standardized Residuals vs. Fitted Values",
    x = "Fitted Values",
    y = "Standardized Residuals",
    color = "Cook's Distance"
  ) +
  theme_minimal()

p1 <- ggplot(fire_linear_fit_aug, aes(x = remoteness, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Mean-Centered Remoteness", y = "Residuals", title = "Residuals vs. Remoteness") +
  theme_minimal()

p2 <- ggplot(fire_linear_fit_aug, aes(x = Wind_pre_15_cent, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Mean-Centered Wind Speed (Pre 30 days)", y = "Residuals", title = "Residuals vs. Wind") +
  theme_minimal()

p3 <- ggplot(fire_linear_fit_aug, aes(x = Temp_pre_15_cent, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Mean-Centered Temperature (Pre 30 days)", y = "Residuals", title = "Residuals vs. Temperature") +
  theme_minimal()

p4 <- ggplot(fire_linear_fit_aug, aes(x = Hum_pre_15_cent, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Humidity (Pre 30 days)", y = "Residuals", title = "Residuals vs. Humidity") +
  theme_minimal()

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Relationship Between Residuals and Predictors",
    subtitle = "", 
    theme = theme(
      plot.title = element_text(size = 14, hjust = 0, family = "sans")
    )
  )
```

To assess whether the normality assumption was reasonably satisfied, we also examined a histogram of residuals. While the distribution of residuals is skewed right, we have far more than 30 observations, so we can assume normality is satisfied.

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

ggplot(fire_linear_fit_aug, aes(x = .resid)) +
  geom_histogram(fill = "cadetblue4", color = "white") +
  labs(
    title = "Histogram of Residuals - Multiple Linear Regression Model",
    x = "Residuals",
    y = "Frequency"
  ) +
  theme_minimal()
```

We also screened for potential collinearity between our predictors and, while we initially observed some concerning multicollinearity between levels of our region variable ($VIF > 5$), we addressed this by collapsing the South and West levels of the variable together and refitting the model based on the recoded version. All of the VIFs for the updated model are less than 5, indicating no issues with multicollinearity among the predictors.

```{r}
vif_values <- vif(fire_linear_fit)

vif_table <- data.frame(
  Predictor = names(vif_values),
  VIF = as.numeric(vif_values)
)

vif_table |>
  arrange(desc(VIF)) |>
  kable(digits = 2, align = "c", col.names = c("Predictor", "VIF")) |>
  kable_styling(full_width = FALSE, font_size = 7)
```

Lastly, we looked for any spatial and temporal correlations in residuals and found no apparent patterns, so independence can be considered reasonably satisfied.

```{r}
#| label: spatial-distribution-residuals
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5


exp_fire_fit_aug <- augment(exp_fire_fit)

fire_linear_fit_aug <- exp_fire_fit_aug |>
  mutate(
    latitude = wildfire_clean$latitude,
    longitude = wildfire_clean$longitude
  )

us_map <- map_data("state")

ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "white", color = "gray60"
  ) +
  geom_point(
    data = fire_linear_fit_aug |> filter(longitude > -130, longitude < -65, latitude > 24, latitude < 50),
    aes(x = longitude, y = latitude, color = .resid),
    size = 1.5, alpha = 0.7
  ) +
  scale_color_distiller(
    palette = "RdBu", direction = 1,
    name = "Residual", limits = c(min(fire_linear_fit_aug$.resid), max(fire_linear_fit_aug$.resid))
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Spatial Distribution of Residuals",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

fire_linear_fit_aug <- fire_linear_fit_aug |>
  mutate(disc_pre_year = wildfire_clean$disc_pre_year)

ggplot(fire_linear_fit_aug, aes(x = disc_pre_year, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "gray40") +
  labs(
    title = "Residuals vs. Year of Discovery",
    x = "Discovery Year",
    y = "Residual"
  ) +
  theme_minimal()
```

#### Methods - Predictive Model

Wildfire records in our dataset can reasonably be treated as independent because, in our assessment of the independence condition for linear regression, we observed no spatial correlations between residuals and geographic location. The randomness assumption was satisfied because the wildfire data were compiled from a large national database covering a wide range of fires across the United States between 1992 and 2015, making it reasonable to assume the dataset is representative of the broader wildfire population. Lastly, the linearity condition appears to be reasonably satisfied, as empirical logit plots for the quantitative predictors did not display strong deviations from linearity, and the assumption of linearity in the log-odds was met.

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

par(mfrow = c(2, 2))

emplogitplot1(mid50_hi ~ Temp_pre_15, data = wildfire_clean, ngroups = 12)
title(main = "Temperature (15 Days Prior)")

emplogitplot1(mid50_hi ~ Hum_pre_15, data = wildfire_clean, ngroups = 12)
title(main = "Relative Humidity (15 Days Prior)")

emplogitplot1(mid50_hi ~ Wind_pre_15, data = wildfire_clean, ngroups = 12)
title(main = "Wind Speed (15 Days Prior)")

emplogitplot1(mid50_hi ~ remoteness, data = wildfire_clean, ngroups = 12)
title(main = "Remoteness (Scaled Distance)")

par(mfrow = c(1, 1))
```

#### Results - Explanatory Model

The estimated multiple linear regression equation predicting log-transformed fire size is shown below:

$$
\begin{aligned}
\widehat{\log(\text{Fire Size})} =\ &\ 2.00 \\
&- 0.94\,\cdot\,\text{Region (Northeast, South/West)} \\
&- 0.26\,\cdot\,\text{Cause Category (Debris, Infrastructure, Other/Unknown, Recreational)} \\
&- 0.31\,\cdot\,\text{Remoteness} \\
&+ 0.19\,\cdot\,\text{Vegetation Group (Grassland, Shrubland, Tundra/Barren)} \\
&+ 0.03\,\cdot\,\text{Wind Speed (15 days prior)} \\
&- 0.00\,\cdot\,\text{Temperature (15 days prior)} \\
&+ 0.01\,\cdot\,\text{Containment Time} \\
&- 0.14\,\cdot\,(\text{Vegetation Group} \times \text{Precipitation (15 days prior)})
\end{aligned}
$$

Additionally, $R^2$ and adjusted $R^2$ values for the fitted model are displayed below:

```{r}
glance(fire_linear_fit) |>
  select(r.squared, adj.r.squared) |>
  rename(
    `R²` = r.squared,
    `Adjusted R²` = adj.r.squared
  ) |>
  kable(digits = 3, format = "latex", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, font_size = 7, latex_options = c("hold_position"))
```

#### Results - Predictive Model

The estimated logistic regression equation predicting the log-odds that a wildfire falls into the upper 50% of moderate-sized fires is shown below.

$$
\begin{aligned}
\log\left(\frac{\hat{\pi}}{1-\hat{\pi}}\right) =\ &\ -0.22 \\
&- 1.83\,\cdot\,\text{Northeast region} \\
&+ 0.18\,\cdot\,\text{South/West region} \\
&- 0.41\,\cdot\,\text{Remoteness} \\
&+ 0.64\,\cdot\,\text{Grassland vegetation} \\
&+ 0.21\,\cdot\,\text{Shrubland vegetation} \\
&+ 0.09\,\cdot\,\text{Tundra/Barren vegetation} \\
&+ 0.03\,\cdot\,\text{Precipitation (15 days prior)} \\
&+ 0.04\,\cdot\,\text{Wind speed (15 days prior)} \\
&- 0.00\,\cdot\,\text{Temperature (15 days prior)} \\
&- 0.00\,\cdot\,\text{Relative humidity (15 days prior)} \\
&- 0.64\,\cdot\,(\text{Grassland vegetation} \times \text{Precipitation}) \\
&- 0.19\,\cdot\,(\text{Shrubland vegetation} \times \text{Precipitation}) \\
&+ 0.01\,\cdot\,(\text{Tundra/Barren vegetation} \times \text{Precipitation})
\end{aligned}
$$

Additionally, the ROC curve for this model is presented below. We observed an AUC of 0.569, indicating relatively limited discriminative ability between fires falling into the upper 50% of moderate-sized fires and those falling into the lower 50%. As we can see in the ROC curve, as the true negative rate increases, sensitivity increases only gradually, which indicates limited improvement in correctly identifying larger fires (even as the true negative rate improves).

```{r}
fire_log_aug <- augment(fire_log_fit)

roc_curve_data <- fire_log_aug |>
  roc_curve(mid50_hi, .fitted, event_level = "second")

autoplot(roc_curve_data)
```

```{r}
fire_log_aug |>
  roc_auc(mid50_hi, .fitted, event_level = "second")
```

Lastly, the results of the drop-in-deviance test for our logistic regression model are presented below. Since the p-value (0.013) was less than 0.05, we retained the interaction between vegetation group and precipitation in our model.

```{r}

fire_log_full <- fire_log_fit

fire_log_reduced <- glm(mid50_hi ~ region + remoteness + 
                   Wind_pre_15 + Temp_pre_15 + Hum_pre_15, data = wildfire_clean,
                   family = "binomial")

anova(fire_log_reduced, fire_log_full, test = "Chisq") |>
  tidy() |>
  mutate(
    term = case_when(
      term == "mid50_hi ~ region + remoteness + Wind_pre_15 + Temp_pre_15 + Hum_pre_15" ~
        "\\makecell{mid50\\_hi ~ region + remoteness + \\\\ Wind\\_pre\\_15 + Temp\\_pre\\_15 + Hum\\_pre\\_15}",
      term == "mid50_hi ~ region + remoteness + Vegetation_group * Prec_pre_15 + Wind_pre_15 + Temp_pre_15 + Hum_pre_15" ~
        "\\makecell{mid50\\_hi ~ region + remoteness + \\\\ Vegetation\\_group * Prec\\_pre\\_15 + \\\\ Wind\\_pre\\_15 + Temp\\_pre\\_15 + Hum\\_pre\\_15}",
      TRUE ~ term
    )
  ) |>
  rename(
    `Model` = term,
    `Residual df` = df.residual,
    `Residual Deviance` = residual.deviance,
    `Test df` = df,
    `Test Deviance` = deviance,
    `p-value` = p.value
  ) |>
  select(Model, `Residual df`, `Residual Deviance`, `Test df`, `Test Deviance`, `p-value`) |>
  kable(digits = 3, format = "latex", booktabs = TRUE, align = "c", escape = FALSE) |> 
  kable_styling(full_width = FALSE, font_size = 7)
```
